---
title: "First steps with the Package jmleIRT"
author: "Jan Steinfeld & Alexander Robitzsch"
date: "`r format(Sys.time(), '%B %d, %Y')`"

output:
  prettydoc::html_pretty:
    smart: false
    theme: cayman
    highlight: github
    css: style.css
    fig_caption: yes
    number_sections: TRUE

vignette: >
  %\VignetteIndexEntry{jmleIRT: Rasch JML (1PL)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
library(jmleIRT)
set.seed(123)
```

## Introduction

This vignette introduces the **jmleIRT** package, which provides efficient joint maximum likelihood estimation (JML) for the Rasch model (1PL) with dichotomous items. 

Key features:

- Tolerates missing responses (`NA`).
- Optional epsilon adjustment (`eps`) to reduce bias from extreme scores.
- Centering based on either "items" or "persons" to ensure identifiability.
- Post-hoc bias correction (`bias_correction = TRUE/FALSE`).
- Optional computation of Warm's weighted likelihood estimates (WLE) for person parameters (`estimatewle = TRUE`).
- Fast Newton-Raphson optimization in C++.

## Simulating Data

We first simulate a dataset of person responses.

```{r simulate-data}
N <- 100  # persons
I <- 10   # items
X <- matrix(rbinom(N * I, 1, 0.5), nrow = N)
# randomly set 10% of entries to missing
X[sample(length(X), size = 0.1 * length(X))] <- NA
head(X)

```

## Basic JML Estimation

Run the JML estimation with centering on "items" (default):

```{r basic-fit}
fit <- jmle_estimation(
  X, max_iter = 500, conv = 1e-5,
  center = "items", bias_correction = FALSE,
  estimatewle = FALSE, verbose = FALSE
)

str(fit)
```

The output is a list with:

- `theta`: person ability estimates
- `beta`: item difficulty estimates
- `iterations`: number of NR iterations
- `converged`: convergence flag
- `bias_correction`: whether applied
- `center`: 'items' or 'persons'  
- `wle_estimate`: person WLEs (if requested)

## Bias Correction

Bias correction rescales item difficulty estimates.

```{r bias-correction}
fit_bc <- jmle_estimation(X, center = "items", bias_correction = TRUE)
summary(fit_bc$beta)
```

Compare to the uncorrected version:

```{r bias-compare}
summary(fit$beta)
```

## Warm's WLE Estimates

Warm’s weighted likelihood estimates can be obtained in addition to JML estimates.

```{r wle-example}
fit_wle <- jmle_estimation(X, center = "items", estimatewle = TRUE)
head(fit_wle$wle_estimate)
```

## Handling Missing Data

The function handles missing responses (`NA`) without failure:

```{r missing-data}
X_miss <- X
X_miss[1:5, 1:2] <- NA
fit_miss <- jmle_estimation(X_miss, center = "items")

length(fit_miss$theta)
length(fit_miss$beta)
```

## Conclusion

The **jmleIRT** package provides a simple and efficient workflow for Rasch JML estimation with useful options such as centering, bias correction, and WLEs, while being robust to missing data.
